R-LIME: LIME法における矩形制約と最適化

### 2024-04-03 paper/paper.tex
ICPR 5th version

\ifnum\mode=0
  \title{%
    R-LIME: LIME法における矩形制約と最適化
  }
  \etitle{%
    R-LIME: Rectangular Constraints and Optimization for Local Interpretable
    Model-agnostic Explanation Methods
  }
  \author{大原 玄嗣 (Genji Ohara)}
  \institute{%
    北海道大学 工学部 \\
    情報エレクトロニクス学科 \\
    情報理工学コース \\
    情報認識学研究室
  }
  \date{2024年3月}
  % \institute{%
  % 	Laboratory for Pattern Recognition and Machine Learning,\\
  % 	Division of Computer Science and Information Technology,\\
  % 	Department of Electronics and Information Engineering,\\
  % 	School of Engineering, Hokkaido University
  % }
\else
  \title{%
    R-LIME: Rectangular Constraints and Optimization for Local Interpretable
    Model-agnostic Explanation Methods}
  % Abbreviated title for the running head
  \titlerunning{%
    R-LIME: Rectangular Constraints and Optimization for LIME Methods
  }

@@

\ifnum\mode=0
  \pagestyle{plain}
  \pagenumbering{roman}
  \tableofcontents
  \newpage
  \listoffigures\listoftables\clearpage
  \AddToHook{cmd/section/before}{\clearpage}
  \pagestyle{plain}
  \pagenumbering{arabic}
\else
  % The abstract should briefly summarize the contents of the paper in
  % 150--250 words.
  \begin{abstract}
    In recent years,
    complex machine learning models have been introduced
    in various industrial fields due to their high accuracy.
    However,
    their complexity has been a major obstacle to implementation
    in sensitive decision-making situations.
    In order to address this problem,
    various post-hoc explanation methods have been proposed,
    but they have not been able to achieve interpretability of
    both the explanation and its scope.
    We propose a new method, R-LIME,
    which interprets complex classifiers in an interpretable scope.
    R-LIME locally approximates a decision boundary of a black-box classifier
    in a rectangular region
    and maximizes the region as long as the accuracy of the linear
    classifier is higher than a given threshold.
    The resulting rectangular region is interpretable for users because it is
    expressed as a conjunction of feature predicates.
    Comparing R-LIME with the existing method,
    we demonstrate the effectiveness of the proposed method through qualitative
    and quantitative experiments using a real-world dataset.
    Finally, we discuss limitations and future work of the proposed method.
    % 161 words
    \keywords{Interpretable machine learning \and Local surrogate model}
  \end{abstract}
\fi

@@

\section{Related Work}
\begin{figure}[tbp]
  \centering
  \def\w{2.2}
  \def\ww{4.4}
  \def\h{0.7}
  \def\hh{1.4}
  \def\hhh{2.1}
  \def\hhhh{2.8}
  \begin{tikzpicture}[scale=0.9,nodestyle/.style={rectangle,node font=\normalsize\itshape,auto}]
    \node[nodestyle] (post-hoc) at (0,0){post-hoc};
    \node[nodestyle] (dependent) at (-\w,-\hh){model-dependent};
    \node[nodestyle] (agnostic) at (\w,-\hh){\textbf{model-agnostic}};
    \node[nodestyle] (global) at (0,-\hhhh){global};
    \node[nodestyle] (local) at (\ww,-\hhhh){\textbf{local}};
    \draw(post-hoc) -- (0,-\h);
    \draw(dependent) -- (-\w,-\h) -- (\w,-\h) -- (agnostic);
    \draw(agnostic) -- (\w,-\hhh);
    \draw(global) -- (0,-\hhh)  -- (\ww,-\hhh) -- (local);
  \end{tikzpicture}
  \caption{%
    Categorization of post-hoc explanation methods.
    We focus on \emph{model-agnostic} and \emph{local} methods,
    which explain model's local behavior using only its output.
  }\label{fig:post-hoc}
\end{figure}
In this chapter,
we overview existing research on post-hoc explanation methods,
which explain the behavior of black-box models already trained.
As shown in \cref{fig:post-hoc},
post-hoc methods are classified into several categories.

First, they are categorized into \emph{model-dependent} and \emph{model-agnostic} methods
based on their dependence on the model's structure.
Model-dependent methods,
such as Deep Taylor Decomposition (DTD)~\cite{montavon2017explaining}
and Layer-wise Relevance Propagation (LRP)~\cite{bach2015pixel},
most of which focus on neural networks and
explain the model's behavior using its parameters~\cite{samek2021explaining}.
While these methods provide detailed explanations
(e.g., layer-wise explanations for neural networks),
it is often challenging
to apply the same method to models with different structures.
On the other hand,
model-agnostic methods use only the output of the model.
Although they are applicable to any model,
they cannot explain the reasoning process inside the model.

@@

\subsection{Previous Work}
We specifically focuses on \emph{local} and \emph{model-agnostic} methods.
In this section,
we briefly review existing research on local model-agnostic explanations,
particularly focusing on studies closely related to our proposed method.

@@

\ifnum\mode=1
  \subsection{Computational Complexity}
  % \subsubsection{Sample Complexity}
  Post-hoc explanation methods including LIME, Anchor, and R-LIME
  need to sample a perturbation vector
  and get the output of the black-box model multiple times,
  which is computationally expensive.
  The number of samples required for LIME is $|\mathcal{Z}_p|$,
  which is the number of samples designated by the user.
  On the other hand,
  the expected number of samples required for Anchor and R-LIME is bounded by
  $\mathcal{O}[m\cdot\mathcal{O}_{\mathrm{MAB}[B\cdot m,B]}]$,
  where $\mathcal{O}_{\mathrm{MAB}[B\cdot m,B]}$
  is the expected number of samples
  for best arm identification finding the best $B$ arms from $B\cdot m$ arms.
  For KL-LUCB algorithm~\cite{kaufmann2013information},
  the expected number of samples is bounded by
  % \begin{equation}
  %   \mathcal{O}\left[H_{\epsilon}\log\frac{H_\epsilon}{\delta}\right],
  % \end{equation}
  % where $H_{\epsilon}$ is sample complexity of the problem
  % \cite{kaufmann2013information}.
  \begin{equation}
    \mathcal{O}_{\mathrm{MAB}[B\cdot m,B]}=
    \mathcal{O}\left[\frac{Bm}{\epsilon^2}\log\frac{Bm}{\epsilon^2\delta}\right].
  \end{equation}

  % \subsubsection{Complexity of Each Trial}
  Each trial of KL-LUCB algorithm consists of sampling a perturbation vector $z$,
  getting output $f(z)$, and updating the linear classifier $g$.
  If we sample a perturbation vector from a multivariate normal distribution
  using Cholesky decomposition in advance ($\mathcal{O}(2^m\cdot m^3)$),
  we can sample $z$ in $\mathcal{O}(m^2)$.
  If we use random forest as the black-box model,
  the complexity of getting output is $\mathcal{O}(1)$.
  If we use logistic regression as the linear classifier and
  update it by stochastic gradient descent (SGD)~\cite{robbins1951stochastic},
  the computational complexity of updating $g$ for each trial is
  $\mathcal{O}(m)$.
  Then the computational complexity of each trial is $\mathcal{O}(m^2)$.

  % \subsubsection{Total Complexity}
  Overall, the complexity of R-LIME is bounded by
  \begin{equation}
    \mathcal{O}\left[
      2^m\cdot m^3+\frac{Bm^4}{\epsilon^2}\log\frac{Bm}{\epsilon^2\delta}
      \right],
  \end{equation}
  which is very sensitive to the number of features $m$.
\fi

@@

A first possible solution to this problem is modifying the loss function.
Using weighted logistic loss or Focal Loss \cite{lin2020focal}
as the loss function might lead to the generation of more useful explanations
in the case of imbalanced label distribution.
Another solution involves adding constraints
to limit the label distribution bias within the approximation region.
In addition to \cref{eq:const_prec}, adding a constraint like
\begin{equation}
  {\left(\mathbb{E}_{z\sim\mathcal{D}(z\mid A)}[\mathbbm{1}_{f(z)=1}]-\frac{1}{2}\right)}^2<\mu
\end{equation}
could suppress the excessive expansion of the approximation region.

@@

We conducted an experiment to evaluate the deviation
between the estimated accuracy and the true accuracy.
We generated explanations for 3,200 data instances sampled from the dataset,
and compared the estimated accuracy with the true accuracy.
The true accuracy was calculated based on 1,000 instances sampled
within the approximation region.
The results in \cref{tab:reward} show a mean deviation of 0.012
with a standard deviation of 0.017.
While the deviation was relatively small considering confidence level $1-\delta=0.95$,
we should modify the selection algorithm to consider the varying accuracy.

\section{Conclusion}
We argued that the interpreting the outputs of black-box models is crucial
for their application in sensitive domains,
and identified challenges in existing methods for local model-agnostic post-hoc
explanations of black-box classifiers.
To address these challenges,
we proposed R-LIME, a method that locally approximates the decision boundary
of a black-box classifier linearly and provides a rectangular approximation region,
which is interpretable for users due to being expressed as a conjunction of feature predicates.
We proposed an algorithm to
maximize coverage of the approximation region
as long as the accuracy of the linear approximation model exceeds a given threshold.
Comparing the outputs of LIME and R-LIME on the real-world dataset,
we demonstrated that R-LIME provides a clear application scope of the
explanation, can be evaluated by users for its reliability and generality,
and achieves higher and less variable local accuracy compared to LIME\@.
However, we discussed the instability of behavior against imbalanced label
distributions and raised questions about the theoretical validity of using
the KL-LUCB algorithm for our problem.

@@

% \subsubsection{Acknowledgements}
% Please place your acknowledgments at
% the end of the paper, preceded by an unnumbered run-in heading (i.e.
% 3rd-level heading).

%
% ---- Bibliography ----
%
% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
%
\ifnum\mode=0
  \bibliographystyle{style/prml}
\else
  \bibliographystyle{style/splncs04}
\fi
\bibliography{myref/myref}
%
\end{document}

>>>>

\ifnum\mode=0
  \title{%
    R-LIME: LIME法における矩形制約と最適化
  }
  \etitle{%
    R-LIME: Rectangular Constraints and Optimization for Local Interpretable
    Model-agnostic Explanation Methods
  }
  \author{大原 玄嗣 (Genji Ohara)}
  \institute{%
    北海道大学 工学部 \\
    情報エレクトロニクス学科 \\
    情報理工学コース \\
    情報認識学研究室
  }
  \date{2024年3月}
  % \institute{%
  %   Laboratory for Pattern Recognition and Machine Learning,\\
  %   Division of Computer Science and Information Technology,\\
  %   Department of Electronics and Information Engineering,\\
  %   School of Engineering, Hokkaido University
  % }
\else
  % The title of the paper
  \title{%
    R-LIME: Rectangular Constraints and Optimization for Local Interpretable
    Model-agnostic Explanation Methods
  }

  % Abbreviated title for the running head
  \titlerunning{%
    R-LIME: Rectangular Constraints and Optimization for LIME Methods
  }

@@

\ifnum\mode=0
  \pagestyle{plain}
  \pagenumbering{roman}
  \tableofcontents
  \newpage
  \listoffigures\listoftables\clearpage
  \AddToHook{cmd/section/before}{\clearpage}
  \pagestyle{plain}
  \pagenumbering{arabic}
\else
  % The abstract should briefly summarize the contents of the paper in
  % 150--250 words.
  \begin{abstract}
    In recent years,
    complex machine learning models have been introduced
    in various industrial fields due to their high accuracy.
    However,
    their complexity has been a major obstacle to implementation
    in sensitive decision-making situations.
    In order to address this problem,
    various post-hoc explanation methods have been proposed,
    but they have not been able to achieve interpretability of
    both the explanation and its scope.
    We propose a new method, R-LIME,
    which interprets complex classifiers in an interpretable scope.
    R-LIME locally approximates a decision boundary of a black-box classifier
    in a rectangular region
    and maximizes the region as long as the accuracy of the linear
    classifier is higher than a given threshold.
    The resulting rectangular region is interpretable for users because it is
    expressed as a conjunction of feature predicates.
    Comparing R-LIME with the existing method,
    we demonstrate the effectiveness of the proposed method through qualitative
    and quantitative experiments using a real-world dataset.
    % 150 words
    \keywords{Interpretable machine learning \and Local surrogate model}
  \end{abstract}
\fi

@@

\section{Related Work}
\begin{figure}[tbp]
  \centering
  \def\w{2.2}
  \def\ww{4.4}
  \def\h{0.7}
  \def\hh{1.4}
  \def\hhh{2.1}
  \def\hhhh{2.8}
  \begin{tikzpicture}[scale=0.9,nodestyle/.style={rectangle,auto}]
    \node[nodestyle] (post-hoc) at (0,0){post-hoc};
    \node[nodestyle] (dependent) at (-\w,-\hh){%
      \begin{tabular}{c}
        model-dependent \\
        (DTD~\cite{montavon2017explaining}, LRP~\cite{bach2015pixel})
      \end{tabular}
    };
    \node[nodestyle] (agnostic) at (\w,-\hh){\textbf{model-agnostic}};
    \node[nodestyle] (global) at (0,-\hhhh){%
      \begin{tabular}{c}
        global \\
        (PDP~\cite{friedman2001greedy}, ALE~\cite{apley2020visualizing})
      \end{tabular}
    };
    \node[nodestyle] (local) at (\ww,-\hhhh){%
      \begin{tabular}{c}
        \textbf{local} \\
        (LIME~\cite{ribeiro2016why}, Anchor~\cite{ribeiro2018anchors})
      \end{tabular}
    };
    \draw(post-hoc) -- (0,-\h);
    \draw(dependent) -- (-\w,-\h) -- (\w,-\h) -- (agnostic);
    \draw(agnostic) -- (\w,-\hhh);
    \draw(global) -- (0,-\hhh)  -- (\ww,-\hhh) -- (local);
  \end{tikzpicture}
  \caption{%
    Categorization of post-hoc explanation methods.
    We focus on \emph{model-agnostic} and \emph{local} methods,
    which explain model's local behavior using only its output.
  }\label{fig:post-hoc}
\end{figure}
In this section,
we overview existing research on post-hoc explanation methods,
which explain the behavior of black-box models already trained.
As shown in \cref{fig:post-hoc},
post-hoc methods are classified into several categories.

They are broadly divided into
\emph{model-dependent} and \emph{model-agnostic} methods
based on their dependence on the model's structure.
Model-dependent methods,
such as Deep Taylor Decomposition (DTD)~\cite{montavon2017explaining}
and Layer-wise Relevance Propagation (LRP)~\cite{bach2015pixel},
most of which focus on neural networks and
explain the model's behavior using its parameters~\cite{samek2021explaining}.
While these methods provide detailed explanations
(e.g., layer-wise explanations for neural networks),
it is often challenging
to apply the same method to models with different structures.
On the other hand,
model-agnostic methods use only the output of the model.
Although they are applicable to any model,
they cannot explain the reasoning process inside the model.

@@

\subsection{Previous Work}
We specifically focus on \emph{local} and \emph{model-agnostic} methods.
In this section,
we briefly review existing research on local model-agnostic explanations,
particularly focusing on studies closely related to our proposed method.

@@

\ifnum\mode=1
  \subsection{Computational Complexity}
  Post-hoc explanation methods including LIME, Anchor, and R-LIME
  need to sample a perturbation vector
  and get the output of the black-box model multiple times,
  which is computationally expensive.
  The number of samples required for LIME is $|\mathcal{Z}_p|$,
  which is the number of samples designated by the user.
  On the other hand,
  the expected number of samples required for Anchor and R-LIME is bounded by
  $\mathcal{O}[m\cdot\mathcal{O}_{\mathrm{MAB}[B\cdot m,B]}]$,
  where $\mathcal{O}_{\mathrm{MAB}[B\cdot m,B]}$
  is the expected number of samples
  for best arm identification finding the best $B$ arms from $B\cdot m$ arms.
  For KL-LUCB algorithm~\cite{kaufmann2013information},
  \begin{equation}
    \mathcal{O}_{\mathrm{MAB}[B\cdot m,B]}=
    \mathcal{O}\left[\frac{Bm}{\epsilon^2}\log\frac{Bm}{\epsilon^2\delta}\right].
  \end{equation}
  Then the total expected number of samples for Anchor and R-LIME is bounded by
  \begin{equation}
    \mathcal{O}\left[\frac{Bm^2}{\epsilon^2}\log\frac{Bm}{\epsilon^2\delta}\right].
  \end{equation}

  For each iteration of KL-LUCB algorithm, R-LIME needs to update
  the linear classifier $g_i$, which is not required in Anchor.
  But if we use logistic regression as the linear classifier,
  the computational complexity of updating $g_i$ is $\mathcal{O}(m)$,
  which is negligible compared to the complexity of sampling a perturbation vector
  and getting the output of the black-box model.
\fi

@@

A first possible solution to this problem is modifying the loss function.
Using weighted logistic loss or Focal Loss \cite{lin2020focal}
as the loss function might lead to the generation of more useful explanations
in the case of imbalanced label distribution.
Another solution involves adding constraints
to limit the label distribution bias within the approximation region.
In addition to \cref{eq:const_prec}, adding a constraint like
\begin{equation}
  {%
    \left(
    \mathbb{E}_{z\sim\mathcal{D}(z\mid A)}[\mathbbm{1}_{f(z)=1}]-\frac{1}{2}
    \right)
  }^2<\mu
\end{equation}
could suppress the excessive expansion of the approximation region.

@@

We conducted an experiment to evaluate the deviation
between the estimated accuracy and the true accuracy.
We generated explanations for 3,200 data instances sampled from the dataset,
and compared the estimated accuracy with the true accuracy.
The true accuracy was calculated based on 1,000 instances sampled
within the approximation region.
The results in \cref{tab:reward} show a mean deviation of 0.012
with a standard deviation of 0.017.
By considering the confidence level $1-\delta=0.95$,
the deviation was relatively small.
While there are concerns about the theoretical validity of using the KL-LUCB algorithm,
the results suggest that the deviation is not significant in practice.

\section{Conclusion}
Existing methods for
local model-agnostic explanations of black-box classifiers have limitations
that they do not explicitly indicate the application scope of the explanations.
To address these challenges,
we proposed R-LIME, a method that locally approximates the decision boundary
of a black-box classifier linearly and provides a rectangular approximation region,
which is interpretable for users due to being expressed as a conjunction of feature predicates.
We proposed an algorithm to
maximize coverage of the approximation region
as long as the accuracy of the linear approximation model exceeds a given threshold.
Comparing the outputs of LIME and R-LIME on the real-world dataset,
we demonstrated that R-LIME provides a clear application scope of the
explanation, can be evaluated by users for its reliability and generality,
and achieves higher and less variable local accuracy compared to LIME\@.
However, we discussed the instability of behavior against imbalanced label
distributions and raised questions about the theoretical validity of using
the KL-LUCB algorithm for our problem.

@@

\ifnum\mode=0
  \bibliographystyle{style/prml}
\else
  \bibliographystyle{style/splncs04}
\fi
\bibliography{myref/myref}

\end{document}
